The challenge provides the data.frame `d` of non tumoral and tumoral tissues described by genes expression values, histological and clinical attributes.

Some gene expression values are missing.

**The goal of the challenge** is to use statistical models (*e.g.* linear models) to predict missing gene expression values using provided gene expression values, histological and clinical attributes.

```{r}
d = readRDS(file="data.rds")
```

 - Biological attributes are:    `r names(d)[c(1:2,4)]`.
 - Histological attribuites are: `r names(d)[c(5,3)]`
 - Clinical attributes are:      `r names(d)[6:10]`
 - Gene columns are named:

```{r, echo=TRUE}
gs = c(
  "ATAD2", "SYCP3",                # Chromatin 
  "H19", "IGF2", "NNAT", "BLCAP",  # inprinted genes
  "BRD4", "BRDT", "NUTM1",         # Testis restricted
  "MAGEB6", "TUBA3C",              # Testis specific
  "SMYD3", "MAP3K2", "KDR",        # K methyl transferase
  "TP53", "KRAS", "BRAF",          # Tumor supressors
  "FASTKD1", "ND2", "ND3", "ND4"   # Mitochondiral activity
)
```


```{r results="verbatim", echo=TRUE}
table(is.na(d[,gs]))
head(d[,gs])
head(d[,1:10])
dim(d)
```

