Please use the `submission_script.Rmd` contained in the `starting_kit` folder as a template to write, run and save your code.


## Generate a missing value prediction

In the provided example 1, we randomly fill missing values.

In the provided example 2, for each gene, we use a simple linear to predict missing values.



```{r echo=TRUE, results="verbnatim", label="generate_zip_file"}
### DO NOT CHANGE THIS PART
d = readRDS("data.rds")
gs = colnames(d)[11:31]

### PUT YOUR SCRIPT HERE
# Example 1
pred = as.matrix(d[,gs])
pred[] = sample(na.omit(pred), nrow(pred)*ncol(pred), replace=TRUE)

# Example 2
pred = sapply(gs, function(g, d){
 m = lm(d[[g]] ~ d$tissue)
#  m = lm(d[[g]] ~ d$age+d$sex+d$tissue+d$tissue_status+d$project)
 predict(m, d)
}, d)

### DO NOT CHANGE THIS PART, it needs *pred* vartiable
pred[!is.na(as.matrix(d[,gs]))] = NA # Sparse, only submit NA values!
saveRDS(pred, "results.rds")
zip_filename = paste(sep="",  "results_", format(Sys.time(), format="%m_%d_%Y_%s"), ".zip")
zip(zip_filename, "results.rds")
print(zip_filename)
```

## Submit zip file

It generates *`r zip_filename`* submission file.

Submit the zip submission file in the `Participate` tab of the codalab challenge by clicking the `Submit` button.

The metric used to evaluate prediction is the mean squared error (MSE) coded with the provided `scoring_function()`.

```{r}
source("scoring_program/scoring.r")
```

On the codalab challenge web page, The *STATUS* become : 
  - Submitting
  - Submitted	
  - Running
  - Finished


When it’s finished, score is given in the column *SCORE* and details for report could be downloaded by clicking *Download output from scoring step*.



